A Recipe for Scavenging - the natural history of a behaviour

### 2016-04-06 natural history of scavenging KH edits.tex
Adam edits to Kev edits

%KH: need paragraph for how do we know something is (was) a scavenger
%TG: Insert figure here: scavenging scale + four points: Locomotion/Detection/Consumption/Environment
\section*{Encounter Rate}
All foraging processes depend on encounter rates between the consumer and the resource. In the simplest case, encounter rates can be thought of in terms of gas diffusion models were the movement of two agents depends on only on the relative speed of the agents. As carcasses are prone to being relatively stationary, the relative speed between a scavenger and carrion is dependent on the movement of the scavenger. As most vertebrates depended on more than simply bumping into resources like gas molecules and can actively detect resources through sensory abilities, scavenging will also depend on search rates which are dependent on both species physiology and the dimensionality of the environment (You guessed it Pawar et al 2012).

\subsection*{Locomotion}
In contrast to the importance of locomotion in predation, locomotion relating to scavenging depends more on the ability to efficiency move over large areas. This ability generally requires an efficient transfer of metabolic energy into movement which may rely on both physiology (i.e. metabolism) and the medium of the environment in which they are moving (aerial, aquatic or terrestrial).

Perhaps the most efficient form of locomotion in vertebrates is paradoxically found in flying species. Despite the energetic costs of flight the only known vertebrate obligate scavengers are the old and the new world vultures. While powered flight is energetically expensive (ref) it is the ability of these species to exploit air currents using large wingspans that allows them to soar while only approximately doubling their metabolic rate \citep{hedenstrom1993migration,spivey2014analysing}. By depending on thermal air flows these species can forage over enormous ranges \citep{spiegel2013factors}. This mode of locomotion is also exploited by sea birds, which use strong ocean winds to search large areas of the oceans \citep{thaxter2012seabird}. While many species of seabird are likely primarily predators, it seems that albatross, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. This is typically in the form of squid, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

The use of soaring in extinct species as a means for scavenging is also likely. The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Mya; ][]{Jetz2012, Jarvis2014} and are  likley to have being soaring flyer's as demonstrated by their large wingspans (Is this true?). (Is there any other form of soaring?). However, soaring flight is likely to be far older than these groups with avian flight originating in the Late Jurassic (163.5-145 Mya) and pterosaurs in the Late Triassic (235-201.3 Mya). Scavenging in the diverse pterosaurs group has been hypothesied many times before \citep{witton2008reappraisal}. Certain clades of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.

While soaring flight is perhaps the only viable means of locomotion that allows for an obligate scavenging life-style (ref Ruxton) powered flight is still an efficient means of locomotion. Flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic} (This needs to be much clearer, needs units etc). We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}. Similarly we would expect that extinct species would also gain in a similar fashion depending on the efficiency of flight. For example, early flying species such as archaeopteryx are predicted to be relatively inefficient (Is there anything on this as I imagine early flight was diverse). Similarly someithing abou t the effecieny of pterosuars flight. The importance of efficient flying over large areas may explain the lack of scavenging behavior in bats as they are general smaller and less efficient in comparison to larger birds (speculating here but I wouldnt be surprised), with only the textit{Necromatis} (``death-eater''), a large bat from the middle to late Eocene (56 to 33.9 Mya) with a robust cranio-mandibular morphology, a likely candidate for scavenging behavior amongst the x amount of known bat species \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}.

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. In fact, the cost of swimming is lower than either running or flying \citep{williams1999evolution} with only soaring flight likely to surpass it (I assume the reference is for powered flight, again this needs to be more precise). This has led some researchers to argue for the feasibility of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. Sharks perhaps best resemble the large soaring flyer's as they depend on large pectoral fins in order to maintain lift. Shark species show very large ranges (for example great white sharks) and are amongst the most lowest cost swimmers (I know the Greenland shark is the slowest swimmer). Faculative scavenging is seen in many marine groups including species of extant sharks such as White sharks (\textit{Carcharodon carcharias}) feeding on whale carcasses \citep{fallows2013white}, Greenland sharks feeding on seals (refs), sixgill sharks (the ref with the pigs) and also extinct species with shark teeth found in the remains of dinosaurs, mososaurs and 4-million-year-old mysticete whales \citep{schwimmer1997scavenging,ehret2009caught}.

Other types of swimming in fish, other species propelled by fins (turtles and plesiosaurs etc)

snakes - ie undulating

endotermy - why it doesn't seem to work
We might expect then that by combining an aquatic environment and an endothermic metabolism that marine mammals would prosper as scavengers.
Fossil pinnipeds and cetaceans from 60 Mya have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, their total energetic cost is similar to an equivalent terrestrial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system.

@@

Terrestrial environments are the most energetically costly in which to move, as a animal must trade off the cost of holding the body up and propelling itself forward. Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}. This is exacerbated by the sprawling gait seen in many lizards which results in the phenomenon known as Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). This would also be true of extinct species. It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
Snakes are a bit different are more effecent at moving I think - laots of scavanging. Both these groups are more lilky to depend on low energetic costs in order to componsate for low ecnounter rates duee to thier inabilty to move. It is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.

@@

%TG: missing some transition here. Maybe something short like "Further back in time" or "Previous to that"...

Something more a bipedal movement.
Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs.
They were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic i.e. intermediate to ecto- and endotherms \citep{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}. %TG: and cite "Kane et al in press" no? I think these people did some kind of work on that matter.
Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles.

Something about humans (also the carnivours kangaroo which might be the most efficient mover.)

>>>>

%KH: need paragraph for how do we know something is (was) a scavenger
%TG: Insert figure here: scavenging scale + four points: Locomotion/Detection/Consumption/Environment
\section*{Encounter Rate}
All foraging processes depend on the encounter rate between consumer and resource. 
In the simplest case, this rate can be thought of in terms of a gas diffusion model where the movement of two agents depends only on their relative speed. As carcasses are stationary, the relative speed between a scavenger and carrion is dependent on the movement of the scavenger. 
Vertebrates do more than simply bump into resources like gas molecules though, because they can actively detect resources through their sensory abilities. 
Scavenging will also depend on search rates, which are dependent on both species physiology and the dimensionality of the environment (\citep{pawar2012dimensionality}).

\subsection*{Locomotion}
In contrast to the importance of locomotion in predation, locomotion relating to scavenging depends more on the ability to efficiently move over large areas. 
This ability generally requires an efficient transfer of metabolic energy into movement which relies on both physiology (i.e. metabolism) and the medium of the environment in which the animal is moving (i.e. aerial, aquatic or terrestrial).
Perhaps the most efficient form of locomotion in vertebrates is, paradoxically, found in flying species. 
Despite the energetic costs of flight, the only known vertebrate, obligate scavengers are the old and the new world vultures. 
This is because, although powered flight is energetically expensive, species like vultures can exploit air currents using large wingspans which allows them to soar while only doubling their metabolic rate \citep{hedenstrom1993migration,spivey2014analysing}.
By depending on thermal air flows these species can forage over enormous ranges \citep{spiegel2013factors}. 
This mode of locomotion is also exploited by sea birds, which use strong ocean winds to search large areas of the oceans \citep{norberg2012vertebrate,thaxter2012seabird}. 
While many species of seabird are likely primarily predators, it seems that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. 
This is typically in the form of squid, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Mya; ][]{Jetz2012, Jarvis2014} % KH and are likley to have being soaring fliers as demonstrated by their large wingspans (Is this true?). (Is there any other form of soaring?).  AK I don't think this is true from what I've seen
However, soaring flight is likely to be far older than this with avian flight originating in the Late Jurassic (163.5-145 Mya) and vertebrate flight in the Late Triassic (235-201.3 Mya) coincident with the pterosaurs. 
Indeed, scavenging in the diverse pterosaurs has been hypothesied many times before \citep{witton2008reappraisal}. 
Certain clades of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
Thus, the use of soaring in extinct species as a means for scavenging is also likely.

While soaring is perhaps the only viable means of locomotion that allows for an obligate, scavenging life-style \citep{ruxton2004obligate} powered flight is still an efficient means of locomotion. 
Avian flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic} 
Even taking account of maintenance costs this still bears out, where total cost of movement (J kg\textsuperscript{-1} m\textsuperscript{-1}) scales according to 5.2 $\times$ body mass (kg) \textsuperscript{-0.23} for fliers and 10.7 $\times$ body mass (kg) \textsuperscript{-0.32} for runners \citep{williams1999evolution}.

We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}. 
Similarly we would expect that extinct species would also gain in a similar fashion depending on the efficiency of flight. 
For example, early birds such as \textit{Archaeopteryx} are predicted to have been poor, relatively inefficient fliers \citep{nudds2010narrow}. 
Similarly someithing about the effecieny of pterosuars flight. 
The importance of efficient flying over large areas may explain the lack of scavenging behavior in bats as they are generally noctural, a time when they would receive no aid from convective air currents \citep{norberg2012vertebrate}. 
%smaller and less efficient in comparison to larger birds (speculating here but I wouldnt be surprised), %AK Don't think it is the case that bats are less efficient than birds in terms of flight 
That said, \textit{Necromatis} (``death-eater''), a large bat from the middle to late Eocene (56 to 33.9 Mya) had a robust cranio-mandibular morphology, and is a likely candidate for scavenging behavior amongst the x amount of known bat species \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}.

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the total cost of movement (again in J kg\textsuperscript{-1} m\textsuperscript{-1}) in fish is lower than either running or flying where it scales according to 2.15 $\times$ body mass (kg) \textsuperscript{-0.25} \citep{williams1999evolution} with only \textit{soaring} flight likely to surpass it. 
%KH(I assume the reference is for powered flight, again this needs to be more precise).  
This has led some researchers to argue for the feasibility of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
Sharks perhaps best resemble the large soaring fliers as they depend on large pectoral fins in order to maintain lift as they swim. 
Many shark species show very large ranges (e.g. great white sharks (\textit{Carcharodon carcharias})) and have the lowest cost of swimming (I know the Greenland shark is the slowest swimmer). 
Faculative scavenging is seen in many marine groups including species of extant sharks such as White sharks, that are known to feed on whale carcasses \citep{fallows2013white}, Greenland sharks feeding on seals \citep{watanabe2012slowest}, and sixgill sharks \citep{anderson2016impact}.
There is evidence too of scavening in extinct species, where shark teeth have been found in the remains of dinosaurs, mososaurs and 4-million-year-old mysticete whales \citep{schwimmer1997scavenging,ehret2009caught}.

% KH Other types of swimming in fish, other species propelled by fins (turtles and plesiosaurs etc)

% KH snakes - ie undulating

% KH endotermy - why it doesn't seem to work
We might expect then that by combining an aquatic environment and an endothermic metabolism marine mammals would prosper as scavengers.
Fossil pinnipeds and cetaceans from 60 Mya have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, the total energetic cost is similar to an equivalent terrestrial or aerial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system.

@@

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move, because an animal must trade off the cost of supporting its body off the ground while propelling itself forward. 
The idea that the evolutionary transition in posture from the sprawling gait of reptiles to the erect posture of mammals conferred a huge advantage to the latter is often repeated e.g. benefits in terms of speed, efficiency, muscle effort and maneuverability \citep{sullivan2015posture}. 
Yet, as \cite{sullivan2015posture} states, most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
Although, metabolic rate unquestionably impacts terrestrial species whereby ectotherms such as many modern reptiles, cannot move for sustained periods \citep{bennett1979endothermy}. 
% Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}. 
This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
This would also be true of extinct species. 
It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
Both these groups are more likely to depend on low energetic costs in order to componsate for low ecnounter rates duee to thier inabilty to move. 
It is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.

@@

%TG: missing some transition here. Maybe something short like "Further back in time" or "Previous to that"...
The evolution of bipedal movment freed up the forelimbs for other purposes e.g. wings, tool-use etc. 
In terms of its cost \cite{alexander2004bipedal} shows that, in the case of humans, we are more economical than predicted while walking and less so while running.

%  KH Something more a bipedal movement.
Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs.
They were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic i.e. intermediate to ecto- and endotherms \citep{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}. %TG: and cite "Kane et al in press" no? I think these people did some kind of work on that matter.
Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles.

% KH Something about humans (also the carnivours kangaroo which might be the most efficient mover.)
